---
title: "expr-CleanUp: preparation and batch correction"
author: "XL"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  rmdformats::readthedown:
    toc_depth: 3
    tabset: true
    number_sections: true
    self_contained: true
    thumbnails: false
    lightbox: false
params:
  batch: APOLLO_COHORT
  forceCorrection: FALSE
  OS: Overall_Survival
  vital_status: Vital_Status
  PFS: Progression-free_survival
  event_for_progression: Event_for_Progression?
  time_point: Timepoint
  best_response: Best_Response_(RECIST)
  best_response_perc: best_response_perc
  Sample_ID: Sample_ID
  MRN: MRN
  mark_library_size: TRUE
  tolerant_library_size_factor: 1.1
  mark_purity: TRUE
  low_purity_threshold: 0.3
  mark_lowexpressgene_pct: TRUE
  lowexpression_threshold: 1
  remove_sample_with_intense_lowexpressgene: TRUE
  outlier_lowexpressgene_pct_factor: 1.5
  remove_lowexpressgene: TRUE
  sample_frequency_threshold: 0.5
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy=TRUE,
  tidy.opts=list(arrow=TRUE, indent=2,width.cutoff=60),
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

require(expr)
require(dplyr)
require(htmltools)
require(plotly)
require(DT)

rm(list = setdiff(ls(),c("params","expressions","sampleAttr")))
w.dir <- getwd()
```


# Prepare clean sample info and expression matrix
```{r load-and-clean-data}
RNA_sample_info=sampleAttr
expressions=expressions

# define class of ID attributes
coerce_ind=names(which(sapply(RNA_sample_info[grep("ID|MRN",colnames(RNA_sample_info))],is.numeric)))
if(length(coerce_ind)>0){RNA_sample_info[,coerce_ind]=as.character(RNA_sample_info[,coerce_ind])}
RNA_sample_info$MRNsurr=MRNsurr(RNA_sample_info$MRN,seed = 123)

# set cleaning param
cleanUpSettings=c("mark_library_size","tolerant_library_size_factor",
                    "mark_purity","low_purity_threshold",
                    "mark_lowexpressgene_pct","lowexpression_threshold",
                    "remove_sample_with_intense_lowexpressgene","outlier_lowexpressgene_pct_factor",
                    "remove_lowexpressgene","sample_frequency_threshold")
paramInput=data.frame(params[cleanUpSettings])

# clean

load(system.file("extdata/protein_coding_ensemble2symbol.RData", package = "expr"))
tmp=expression(do.call(
  prepare_clean_RNA_sample_info_and_protein_expressions,
  c(list(
    RNA_sample_info=RNA_sample_info,
    expressions = expressions,
    sample_id_col = "Sample_ID",gene_id_col = "rowname",
    adjusted_by_library_size = T,ens_id_col = "Gene_ID",symbol_col = "Gene_Symbol",
    protein_coding_ensemble2symbol_table = protein_coding_ensemble2symbol),
    as.list(paramInput)
  )))

consoleOutput=capture.output(results<-eval(tmp),type = c("output"))


# data loading summary
SumInfo=t(data.frame(
  `Number_of_Cleaned_Samples`=dim(results$clean_RNA_sample_info)[1],
  `Number_of_Cleaned_Genes`=dim(results$clean_protein_expressions)[1],
  `Number_of_Deleted_Samples`=dim(results$marked_ori_RNA_sample_info)[1]-dim(results$clean_RNA_sample_info)[1],
  paramInput[,grep("factor|threshold",colnames(paramInput))],
  Notes=capture.output(grep("too many low expression genes",consoleOutput,value = T),type = c("output"))))# when no samples were deleted
t1=kableExtra::kbl(SumInfo,caption = "Data loading summary")%>%kableExtra::kable_styling(position = "center",font_size = "200%")


# cleaned RNA sample info
# remove foundry ID columns enclosing matrix or vector; remove patient ID
t2=results$clean_RNA_sample_info
t2=t2[,!colnames(t2) %in% c("protocol_numbers","PID","Core_ID","MRN","Patient_ID","full",grep("file",colnames(t2),value = T))]
t2.0=t2=t2[gtools::mixedorder(t2$MRNsurr),] %>% 
  tibble::remove_rownames() %>% tibble::column_to_rownames("Sample_ID") %>% tibble::rownames_to_column("Sample_ID")
t2=format(t2,nsmall=2) %>%
  DT::datatable(options = list(
    # scrollX=T,scrollY=T,
    initComplete = JS("function(settings, json) {","$(this.api().table().header()).css({'font-size': '12px'});","}")
    ),
    rownames = F,
    # caption = htmltools::tags$caption(
    #   style = 'caption-side: top; text-align: left; color:black; font-size:200% ;',
    #   'Sample Info (cleaned)') 
    ) %>%
    formatStyle(columns = colnames(.$x$data), `font-size` = '12px')


# log2 transformation
clean_RNA_sample_info<-results[["clean_RNA_sample_info"]]
clean_protein_expressions<-results[["clean_protein_expressions"]]
log2_clean_protein_expressions<-log2(clean_protein_expressions+1)

```

```{r load-and-clean-data-res, echo=FALSE, results = 'asis'}
cat(paste0("## ","Data loading summary."), sep = "\n")
t1

cat(paste0("\n## ","Cleaned RNA sample info."), sep = "\n")
t2
```

# Batch effects
## Visualization of sample distribution via unsupervised clustering
In this step, genes with higher variance across data set and a fill rate greater than 0.75 are used for unsupervised analysis.
```{r data-load}

rm(list = setdiff(ls(),c(lsf.str(), grep("clean|protein_coding_gene_info|0|params",ls(),value = T))))

# set param
paramInput=data.frame(method="MAD",mad_top_n = 1000,remove_outlier = F)

# prep data
unsupervised_data<-prepare_unsupervised_data(
  expressions=log2_clean_protein_expressions,
  method=paramInput[["method"]],
  mad_top_n = paramInput[["mad_top_n"]],
  remove_outlier = paramInput[["remove_outlier"]])
# assertthat::are_equal(colnames(unsupervised_data),clean_RNA_sample_info[["Sample_ID"]])


```

```{r batch-effect-visualization, fig.show='hide'}

# run analysis
  umap.config=umap::umap.defaults
  umap.config$n_neighbors=round(ncol(unsupervised_data)/2)

labels=as.factor(clean_RNA_sample_info$batch)
unsupervised_results=unsupervised_analysis(unsupervised_data,labels = labels,umap_config = umap.config,
                                           cols = setNames(cols4all::c4a("dark24",nlevels(labels)),levels(labels)))

# generate plots
p1=unsupervised_results$plots$pca$`3D`%>%layout(title = "PCA", autosize = F)

p2=ggplotly(
  unsupervised_results$plots$umap +
    theme_bw() + theme(legend.position='none') +
    geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
  tooltip = c("x","y","Label","text")) %>%
    layout(title = "UMAP")

p3=ggplotly(
  unsupervised_results$plots$tsne +
    theme_bw() + theme(legend.position='none') +
    geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
  tooltip = c("x","y","Label","text")) %>%
  layout(title = "tSNE")

p4=ggplotly(
  unsupervised_results$plots$mds +
    theme_bw()+ theme(legend.position='none') +
    geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
  tooltip = c("x","y","Label","text"))%>%
  layout(title = "MDS")

p5=unsupervised_results$plots$heatmap
```

```{r batch-effect-visualization-plots,class.source = 'fold-hide'}
a=data.frame(x=c(0.2,0.8,0.2,0.8),
           y=c(1.0,1.0,0.45,0.45),
           text=c("PCA","UMAP","tSNE","MDS"),
           xref = "paper",
           yref = "paper",
           xanchor = "center",
           yanchor = "bottom",
           showarrow = FALSE 
           )
annotations = apply(a,1, function(x) lapply(as.list(x),type.convert, as.is=TRUE))

# grid.draw(cowplot::get_legend(unsupervised_results$plots$umap+theme(legend.position = "top")))
p=subplot(list(style(p1,showlegend = F),p2,p3,p4),nrows = 2,margin = 0.1) %>%
  layout(title = "Unsupervised clustering before correction",
         annotations = annotations,
         scene = list(domain=list(x=c(0,0.5),y=c(0.5,1))))

browsable(tagList(tags$div(
  style = 'width:100%;height=12;display:block;float:left;',
  p)))
```

```{r batch-effect-visualization-plots2,results='asis',fig.width=6,fig.height=3,out.width="100%",dpi=130,class.source = 'fold-hide'}
cat("<br>")
p5
```

## Determine if batch effect exist
Call Batch effect "exist", if k-means clustering match with batch definition, and that clusters are clearly separated ( evaluated by Silhouette score and withinss/betweenss). If **batch_effect_exist==TRUE**, proceed to correction; Otherwise, results in next section stays blank.
```{r batch-effect-determination, results = 'asis'}
kmean.df=(unsupervised_results$ress$umap$layout)
# calculate Silhouette score for k from 2 to 2 times number of batches and determine the optimal k for clustering
avg_sil <- function(k) {
  km.res <- kmeans(kmean.df, centers = k, nstart = 25)
  ss <- cluster::silhouette(km.res$cluster, dist(kmean.df))
  mean(ss[, 3])}
k.values = 2:(2*(length(unique(clean_RNA_sample_info$batch))))
avg_sil_values <- sapply(k.values, avg_sil)
km.res <- kmeans(kmean.df, centers = k.values[which.max(avg_sil_values)], nstart = 25)

# test if good clustering
good=all(c(km.res$betweenss/km.res$totss>0.5,
      max(avg_sil_values)>0.1))

# test if k-means clustering match with batch definition
clusters=lapply(unique(km.res$cluster), function(x) names(km.res$cluster)[km.res$cluster==x])
batches=lapply(unique(clean_RNA_sample_info$batch),
               function(x) clean_RNA_sample_info$Sample_ID[clean_RNA_sample_info$batch==x])
hits=c()
i=0
for (ncluster in 1:length(clusters)){
  for (nbatches in 1:length(batches)){
    hit=length(setdiff(clusters[[ncluster]],batches[[nbatches]]))==0 &
      length(setdiff(batches[[nbatches]],clusters[[ncluster]]))==0
    hits=c(hits,hit)
  }
}
matched=any(hits)==T

batch_effect_exist=all(matched,good)
cat(sprintf("Clustering good?: **%s**<br>
            Clustering matched?: **%s**<br>
            Batch effect exists: **%s**<br>
            Force correction: **%s**<br>",good,matched,batch_effect_exist,params$forceCorrection))
batch_effect_exist=batch_effect_exist|params$forceCorrection
```

## Batch effect correction
If batch effect exist, attempt to use `limma` and `COMBAT` to correct such effect
```{r batch-effect-correction}
require(umap)
  umap.config=umap::umap.defaults
  umap.config$n_neighbors=round(ncol(log2_clean_protein_expressions)/2)
  unsupervised_resultss=list()
if(batch_effect_exist){
  #limma
  limma_rbe_log2_protein_expressions<-limma::removeBatchEffect(
    log2_clean_protein_expressions,batch=clean_RNA_sample_info$batch)
  uns_limma_rbe_log2_protein_expressions<-prepare_unsupervised_data(limma_rbe_log2_protein_expressions,method="MAD",mad_top_n=1000)
  labels=as.factor(clean_RNA_sample_info$batch)
  unsupervised_resultss[["limma"]]=unsupervised_analysis(uns_limma_rbe_log2_protein_expressions,labels=labels,umap_config = umap.config,cols = setNames(cols4all::c4a("dark24",nlevels(labels)),levels(labels)))
  #combat
  combat_rbe_log2_protein_expressions<-sva::ComBat(log2_clean_protein_expressions,batch=clean_RNA_sample_info$batch)
  uns_combat_rbe_log2_protein_expressions<-prepare_unsupervised_data(combat_rbe_log2_protein_expressions,method="MAD",mad_top_n=1000)
  unsupervised_resultss[["combat"]]=unsupervised_analysis(uns_combat_rbe_log2_protein_expressions,labels=labels,umap_config = umap.config,cols = setNames(cols4all::c4a("dark24",nlevels(labels)),levels(labels)))


# generate plots
p1=p2=p3=p4=p5=list()
for (method in c("limma","combat")){
  unsupervised_results=unsupervised_resultss[[method]]
  p1[[method]]=unsupervised_results$plots$pca$`3D`%>%layout(title = sprintf("PCA %s",method), autosize = F)
  
  p2[[method]]=ggplotly(
  unsupervised_results$plots$umap +
    theme_bw() + theme(legend.position='none') +
    geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
  tooltip = c("x","y","Label","text")) %>%
    layout(title = sprintf("UMAP %s",method))

  p3[[method]]=ggplotly(
    unsupervised_results$plots$tsne +
      theme_bw() + theme(legend.position='none') +
      geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
    tooltip = c("x","y","Label","text")) %>%
    layout(title = sprintf("tSNE %s",method))
  
  p4[[method]]=ggplotly(
    unsupervised_results$plots$mds +
      theme_bw()+ theme(legend.position='none') +
      geom_point(aes(text=sprintf("sample: %s", colnames(unsupervised_data)))),
    tooltip = c("x","y","Label","text"))%>%
    layout(title = sprintf("MDS %s",method))
  
  p5[[method]]=unsupervised_results$plots$heatmap
}
}
```

### Correction result {.tabset}
#### limma result
```{r limma-plots,class.source = 'fold-hide'}
if (batch_effect_exist) {
a=data.frame(x=c(0.2,0.8,0.2,0.8),
           y=c(1.0,1.0,0.45,0.45),
           text=c("PCA","UMAP","tSNE","MDS"),
           xref = "paper",
           yref = "paper",
           xanchor = "center",
           yanchor = "bottom",
           showarrow = FALSE 
           )
annotations = apply(a,1, function(x) lapply(as.list(x),type.convert, as.is=TRUE))

pp1=p1[["limma"]]
pp2=p2[["limma"]]
pp3=p3[["limma"]]
pp4=p4[["limma"]]
p=subplot(list(style(pp1,showlegend = F),pp2,pp3,pp4),nrows = 2,margin = 0.1) %>%
  layout(title = "Unsupervised clustering after correction",
         annotations = annotations,
         scene = list(domain=list(x=c(0,0.5),y=c(0.5,1))))

browsable(tagList(tags$div(
  style = 'width:100%;height=12;display:block;float:left;',
  p)))
}
```

```{r limma-plots2,results='asis',fig.width=6,fig.height=3,out.width="100%",dpi=130,class.source = 'fold-hide'}
if (batch_effect_exist) {
cat("<br>")
p5[["limma"]]
}
```

#### COMBAT result
```{r combat-plots,class.source = 'fold-hide'}
if (batch_effect_exist) {
a=data.frame(x=c(0.2,0.8,0.2,0.8),
           y=c(1.0,1.0,0.45,0.45),
           text=c("PCA","UMAP","tSNE","MDS"),
           xref = "paper",
           yref = "paper",
           xanchor = "center",
           yanchor = "bottom",
           showarrow = FALSE 
           )
annotations = apply(a,1, function(x) lapply(as.list(x),type.convert, as.is=TRUE))

pp1=p1[["combat"]]
pp2=p2[["combat"]]
pp3=p3[["combat"]]
pp4=p4[["combat"]]
p=subplot(list(style(pp1,showlegend = F),pp2,pp3,pp4),nrows = 2,margin = 0.1) %>%
  layout(title = "Unsupervised clustering after correction",
         annotations = annotations,
         scene = list(domain=list(x=c(0,0.5),y=c(0.5,1))))

browsable(tagList(tags$div(
  style = 'width:100%;height=12;display:block;float:left;',
  p)))
}
```

```{r combat-plots2,fig.width=6,fig.height=3,out.width="100%",dpi=130,class.source = 'fold-hide'}
if (batch_effect_exist) {
cat("<br>")
p5[["combat"]]
}
```

```{r save}
if (batch_effect_exist) {
  save(batch_effect_exist,log2_clean_protein_expressions,limma_rbe_log2_protein_expressions,combat_rbe_log2_protein_expressions,clean_RNA_sample_info,file = "clean_batchexamined_logRNA.RData")
}else{
  save(batch_effect_exist,log2_clean_protein_expressions,clean_RNA_sample_info,file = "clean_batchexamined_logRNA.RData")
}
rm(list=ls())
```
